---
title: "project"
author: "Varshini Vaisnavi Srinivasan"
date: "2024-03-30"
output: word_document
---


```{r}
#installing packages
#install.packages(c('lubridate', 'imputeTS', 'urca', 'lmtest', 'TTR', 'dplyr',
 #                 'forecast', 'pastecs', 'seastests', 'tseries', 'uroot', 'crayon', 'Hmisc'))

```
```{r}

options(repos= 'https://cran.rstudio.com/')
# --- Importing Libraries ---
suppressPackageStartupMessages({
  library(lubridate)
  library(imputeTS)
  library(urca)
  library(lmtest)
  library(TTR)
  library(dplyr)
  library(forecast)
  library(pastecs)
  library(seastests)
  library(tseries)
  library(uroot)
  library(crayon)
  library(ggplot2)
})
```

```{r}
# Libraries Settings
options(warn = -1)
options(repr.plot.width = 14, repr.plot.height = 10)

#Importing Dataset
data <- read.csv('crude oil.csv')

#subset
data <- data %>% select(LOCATION, TIME, Value) 

oil_canada <- data %>% filter(LOCATION == 'CAN', TIME > 1970)
oil_german <- data %>% filter(LOCATION == 'DEU', TIME > 1970)


cat(blue$bold('Dataset:\n'))
cat(black('Canada: '))
head(oil_canada, 5)
cat(black('German: '))
head(oil_german, 5)

#Dataset Info
cat(blue$bold('Dataset Info:\n'))

cat(black('Canada :\n'))
cat(reset('\n\t Total Rows: ',dim(oil_canada)[1]))

cat(reset('\n\t Total Columns: ',dim(oil_canada)[2]))
cat(black('German: \n'))
cat(reset('\n\tTotal Rows: ',dim(oil_german)[1]))
cat(reset('\n\t Total Columns: ',(oil_german)[2]))

```

***EDA***
```{r}
cat(black('Canada: '))
oil_canada %>% summarise(LOCATION = sum(is.na(LOCATION) == 1), 
                      TIME = sum(is.na(TIME) == 1),
                      Value=sum(is.na(Value) == 1))

cat(black('German: '))
oil_german %>% summarise(LOCATION = sum(is.na(LOCATION) == 1), 
                      TIME = sum(is.na(TIME) == 1),
                      Value=sum(is.na(Value) == 1))


```

***Time Series Plot***

```{r}
#New Dataframe
new_can = ts(oil_canada$Value, frequency = 1, start = 1971)
```
```{r}
ts_ger = ts(oil_german$Value, frequency = 1, start = 1971)
```
```{r}
#Time Series Plot
autoplot(new_can, series = 'Canada', lwd = 1.2) +
  autolayer(ts_ger, series = 'German', lwd = 1.2)+
  scale_color_manual(values = c('#6F1D1B', '#DBB42C'))+
  labs(title = 'Crude Oil Value Plot', x = '\nYear', y = 'Value\n', 
       subtitle = 'in Canada and German')+
  geom_hline(yintercept = 91517.88, linetype = 'longdash', color = '#800000')+
  annotate('text', x = 2017, y = 85000, label = 'Avg. Value:\n91517.88', size = 2, color = '#290100')+
  annotate('text', x = 1986, y = 100000, label = '1982 - Lowest crude oil value\n(57339.49)', size = 2, fontface = 'italic', 
           color = '#290100')+
  annotate(geom = 'curve', x = 1984, y = 90000, xend = 1982, yend = 60000, curvature = 0.1, arrow = arrow(length = unit(0.5, 'cm')),
           color = '#290100')+
  annotate('text', x = 2004, y = 137500, label = 'Crude oil value continues to\nincrease in the past 8 years', size = 2, fontface = 'italic', 
           color = '#290100')+
  annotate(geom = 'curve', x = 2006, y = 130000, xend = 2009, yend = 110000, curvature = -0.2, arrow = arrow(length = unit(0.5, 'cm')),color = '#290100')+
  annotate('text', x = 2011, y = 162500, label = '2017 - Highest crude\noil value (174226.07)', size = 2, fontface = 'italic', 
           color = '#290100')+
  annotate(geom = 'curve', x = 2013, y = 170000, xend = 2016, yend = 174300, curvature = -0.1, arrow = arrow(length = unit(0.5, 'cm')),
           color = '#290100')+
  geom_hline(yintercept = 3898.02, linetype = 'longdash', color = '#B69121')+
  annotate('text', x = 2017, y = 11000, label = 'Avg. Value:\n3898.02', size = 2, color = '#76520E')+
  annotate('text', x = 2000, y = 15000, label = 'Crude oil value continues to\ndecline until 2017', size = 2, fontface = 'italic', 
           color = '#76520E')+
  theme(panel.background = element_rect(fill = 'white', colour = 'white'),
        plot.background = element_rect(fill = 'white', colour = 'white'),
        plot.title = element_text(size = 14, colour = 'gray15', face = 'bold', hjust = 0.5, vjust = 0.5),
        plot.subtitle = element_text(size = 12, colour = 'gray15', face = 'italic', hjust = 0.5, vjust = 0.5),
        plot.caption = element_text(color = 'gray25', face = 'italic', size = 12),
        panel.grid.major.y = element_line(linetype = 'dotted', colour = 'gray70'),
        panel.grid.minor.y = element_line(linetype = 'dotted', colour = 'gray75'),
        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),  
        axis.title = element_text(size = 14, face = 'bold', colour = 'gray15'),
        axis.text = element_text(size = 12, colour = 'gray8'), legend.position = 'bottom',
        axis.line = element_line(), legend.text = element_text(size = 10),
        legend.title = element_text(colour = 'gray5', size = 14, face = 'bold'),
        legend.background = element_rect(fill = 'white', linetype = 'blank'))
```

* Canada: - upward trend with no seasonality
          -  non- stationary

* German: - downward trend with no seasonality
          - stationary


***Statistical and Non Statistical Test***

Test Hypothesis:

H0: no trend available 

H1: trend available

if p-value is less than 0.05, reject H0

```{r}

# Mann-Kendall Trend Test
cat(blue$bold('Mann-Kendall Trend Test :.\n'))

cat(black('Canada '))
trend.test(new_can)

cat(black('German '))
trend.test(ts_ger)

```
Canada:
* p-value for Canada is less than 0.05, so no trend; H0 accepted

German: 
* p-value for Germany is less than 0.05, so no trend; H0 accepted

**ACF and PACF Plot**

```{r}
# --- ACF & PACF Plot (Canada) ---
tsdisplay(new_can, main = 'ACF and PACF Plot - Canada', col = '#F00000', lwd = 1.5, xlab = 'Year', ylab = 'Oil Value', cex.main = 1.5, cex.lab = 1.15)
title(adj = 1, cex.sub = 1, font.sub = 3)
```

*series for Canada is not stationary, autocorrelation remains significant for the first several lags

*ACF dies down exceptionally slowly

```{r}
#ACF & PACF Plot
tsdisplay(ts_ger, main = 'ACF and PACF Plot - German', col = '#F00000', lwd = 1.5, xlab = 'Year', ylab = 'Oil Value', cex.main = 1.5, cex.lab = 1.15)
title(adj = 1, cex.sub = 1, font.sub = 3)
```

* German is not stationary ,remain significant for first several lags

* ACF plot dies down exceptionally slowly

***Unit Root Test***

H0: not stationary
H1: stationary
```{r}
#ADF Test 
cat(blue$bold('ADF Test:\n'))

cat(black$bold('Canada: '))
adf.test(new_can)

cat(black$bold('German: '))
adf.test(ts_ger)
```
Canada:

* p-value for Canada is greater than 0.05, not stationary

German:

* p-value for German is greater than 0.05, not stationary


***Forecasting Techniques***

```{r}
#Splitting Dataframe
split_can = round(length(new_can)*0.7, 0)
split_ger = round(length(ts_ger)*0.7, 0)

#Canada - Train & Test
train_can = window(new_can, end = 2003)
test_can = window(new_can, start = 2004)

#German - Train & Test
train_ger = window(ts_ger, end = 2003)
test_ger = window(ts_ger, start = 2004)

#Train & Test Detail
cat(blue$bold('.: Train-Test Info:\n'))
cat(black$bold('Canada: \n'))
cat(reset('\tTrain size: '))
cat(bold(length(train_can), '\n'))
cat(reset('\tTest size: '))
cat(bold(length(test_can), '\n'))
cat(black$bold('German: \n'))
cat(reset('\tTrain size: '))
cat(bold(length(train_ger), '\n'))
cat(reset('\tTest size: '))
cat(bold(length(test_ger), '\n'))
```

**SES**

alpha is called smoothing parameter, values between 0 to 1

α =0, future forecasted values based on the average of past data
α = 1, future forecasted values based on current observation 

```{r}
#Applying SES
cat(black$bold('Canada : \n'))
fitses_canada = ses(train_can, alpha = 0.8, initial = 'simple')
plot(forecast(fitses_canada, data.frame(test_can, h = length(test_can))), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(fitses_canada), col = '#78380C', lwd = 2.5)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)

title(main = 'Forecast Results - Canada', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 1)
title(main = 'using Simple Exponential Smoothing', cex.main = 1, font.main = 2, line = 0.2)

# Summary
cat(black$bold('.: SES Summary :.'))
summary(fitses_canada)
cat(reset('\n'))
```
```{r}
cat(black$bold('Canada : \n'))
accuracy(fitses_canada, test_can)
```
```{r}
#Applying SES
cat(black$bold('German : \n'))
fitses_ger = ses(train_ger, alpha = 0.8, initial = 'simple')
plot(forecast(fitses_ger, data.frame(test_ger, h = length(test_ger))), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(fitses_ger), col = '#C32F27', lwd = 2.5)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - German', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using Simple Exponential Smoothing', cex.main = 1, font.main = 3, line = 0.2)
title(adj = 1, cex.sub = 1, font.sub = 3)

cat(blue$bold('.: SES Summary :.'))
summary(fitses_ger)
cat(reset('\n'))
```
```{r}
cat(black$bold('German : \n'))
accuracy(fitses_ger, test_ger)
```
```{r}
#ETS
cat(black$bold('Canada : \n'))
ets_canada = ets(train_can)
ets_canada_forecast = ets(test_can, model = ets_canada, use.initial.values = FALSE)
plot(forecast(ets_canada, h = 5), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(ets_canada), col = '#503A2B', lwd = 2.5)

grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - Canada', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using ETS', cex.main = 1, font.main = 3, line = 0.2)
title(adj = 1, cex.sub = 1, font.sub = 3)

cat(bold('.: ETS Summary :.'))
summary(ets_canada)
cat(reset('\n'))
```


```{r}
cat(blue$bold('German : \n'))

ets_ger = ets(train_ger)
ets_ger_forecast = ets(test_ger, model = ets_ger, use.initial.values = FALSE)
plot(forecast(ets_ger, h = 5), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(ets_ger), col = '#FFBA08', lwd = 2.5)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - German', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using ETS', cex.main = 1, font.main = 3, line = 0.2)
title(adj = 1, cex.sub = 1, font.sub = 3)

cat(black$bold('.: ETS Summary :.'))
summary(ets_ger)
cat(reset('\n'))
```


```{r}

cat(blue$bold('Canada : \n'))

naive_canada = naive(train_can)
plot(forecast(naive_canada, h = 5), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(naive_canada), col = '#A94923', lwd = 2.5)

#Plot 
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)

title(main = 'Forecast Results - Canada', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using Moving Average', cex.main = 1, font.main = 3, line = 0.2)
title(adj = 1, cex.sub = 1, font.sub = 3)

cat(black$bold('.: MA Summary :.'))
summary(naive_canada)
cat(reset('\n'))
```

```{r}
cat(blue$bold('German : \n'))
naive_ger = naive(train_ger)
plot(forecast(naive_ger, h = 5), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(naive_ger), col = '#FFDD00', lwd = 2)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)

title(main = 'Forecast Results - German', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using Moving Average', cex.main = 1, font.main = 3, line = 0.2)
title(adj = 1, cex.sub = 1, font.sub = 3)

cat(black$bold('.: MA Summary :.'))
summary(naive_ger)
cat(reset('\n'))
```

```{r}

#Comparison Accuracy
cat(blue$bold('Canada : \n'))

#SES Accuracy
cat(black$bold('\n.: SES Accuracy :.\n'))
accuracy(fitses_canada, test_can)

#ETS Accuracy
cat(black$bold('.: ETS Accuracy :.\n'))
ets_canada_acc = rbind(accuracy(ets_canada), accuracy(ets_canada_forecast))
rownames(ets_canada_acc) <- c('Training Set', 'Test Set')
ets_canada_acc

#MA Accuracy
cat(black$bold('\n.: Moving Average Accuracy :.\n'))
accuracy(naive_canada, test_can)

```
```{r}

cat(blue$bold('Germany : \n'))

#SES Accuracy
cat(black$bold('\n.: SES Accuracy :.\n'))
accuracy(fitses_ger, test_ger)

#ETS Accuracy
cat(black$bold('\n.: ETS Accuracy :.\n\n'))
ets_ger_acc = rbind(accuracy(ets_ger), accuracy(ets_ger_forecast))
rownames(ets_ger_acc) <- c('Training Set', 'Test Set')
ets_ger_acc

#Naive Model Accuracy
cat(black$bold('\n.: Moving Average Accuracy :.\n\n'))
accuracy(naive_ger, test_ger)

```

Canada: 

* In general, models are overfitting, since RMSE and MAE values for test is higher compare to training

* From MAPE results, ETS model provides an error percentage of 5.9% in training and 5.1% in testing

German:

* In general, models are overfitting, since RMSE and MAE values for test higher compare to training

* From MAPE results, ETS model error percentage of 2.9% for training and 5.6% for testing



***Non- Seasonal Differencing***

*Non-seasonal differencing is needed for both countries in order to get stationary time series 

*perform non-seasonal differencing before applying suitable ARIMA model


**ACF and PACF Plot after 2 Non-Seasonal Differencing**
```{r}
#Canada - ACF and PACF after 2 Non-Seasonal Differencing
tsdisplay(diff(diff(new_can, 1), 1), main = 'Canada - ACF & PACF after 2 Differencing', col = '#844F39', lwd = 1.5, xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15)

```

```{r}

#German - ACF and PACF after 2 Non-Seasonal Differencing
tsdisplay(diff(diff(ts_ger,1),1), main = 'German - ACF & PACF after 2 Differencing', col = '#DC5713', lwd = 1.5, xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15)

```


**ARIMA Models**

ARIMA models that will be built will be defined based on ACF and PACF plot for both countries. 

Canada:

* ACF: cut off after lag 1/ MA(1)

* PACF: cut off after lag 1/ AR(1)

* Differencing needed (d) = 2

Conclusion: ARIMA (1,2,1)

German:

ACF: cut off after lag 1/ MA(1)

PACF: cut off after lag 1/ AR(1)

Differencing needed (d) = 2

Conclusion: ARIMA(1,2,1)

Comparsion purpose, ARIMA(1,2,2) will be both countries

**ARIMA(1,2,1):**

```{r}
#Canada - ARIMA(1,2,1)
arima_can_121 = Arima(new_can, order = c(1,2,1))

#German - ARIMA(1,2,1)
arima_ger_121 = Arima(ts_ger, order = c(1,2,1))

```
**ARIMA(1,2,2):**

```{r}
#Canada - ARIMA(1,2,2)
arima_can_122 = Arima(new_can, order = c(1,2,2))

#German - ARIMA(1,2,2)
arima_ger_122 = Arima(ts_ger, order = c(1,2,2))
```

***ARIMA Models Comparsion***

**Relative Performance (AIC, AICc and BIC)**

Lower the AIC, AICc and BIC values the better ARIMA model

```{r}
#Canada - ARIMA Summary
cat(blue$bold('Canada : \n'))

cat(black$bold('\n.: ARIMA 1 Accuracy :.\n'))
summary(arima_can_121)
cat(black$bold('\n.: ARIMA 2 Accuracy :.\n'))
summary(arima_can_122)
```

It can be concluded that model ARIMA(1,2,1) has better AIC, AICc, BIC values

```{r}
#German - ARIMA Summary
cat(blue$bold('German : \n'))
cat(black$bold('.: ARIMA 1 Accuracy :.'))
summary(arima_ger_121)
cat(black$bold('.: ARIMA 2 Accuracy :.'))
summary(arima_ger_122)
```
It can concluded that the model ARIMA(1,2,1) has better AIC, AICc and BIC values


**Forecasting using ARIMA(1,2,1)**

Forecasting for the following ten periods for both countries

```{r}

#Plot
plot(forecast(arima_can_121, h = 10), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(arima_can_121), col = 'goldenrod2', lwd = 2.5)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - Canada', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using Best ARIMA Model - ARIMA(1,2,1)', cex.main = 1, font.main = 3, line = 0.2)


#Forecast
cat(black$bold('.: Canada - Forecast Results (Best ARIMA) :.\n'))
forecast(arima_can_121, h = 10)
cat(reset('\n'))
```

```{r}
#Plot
plot(forecast(arima_ger_121, h = 10), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(arima_ger_121), col = 'red2', lwd = 2.5)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - German', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using Best ARIMA Model - ARIMA(1,2,1)', cex.main = 1, font.main = 3, line = 0.2)


#Forecast 
cat(black$bold('.: German - Forecast Results (Best ARIMA) :.\n'))
cat(reset$black('*********************************************'))
forecast(arima_ger_121, h = 10)
cat(reset('\n'))
```

***SARIMA***
```{r}
library(astsa)

difflog_can<-diff(diff(log(new_can)))

difflog_ger<-diff(diff(log(ts_ger)))

```
```{r}

sarima_121_can <- arima(difflog_can, order = c(1, 2, 1), seasonal = list(order = c(1, 1, 0), period = 12))

sarima_121_ger <- arima(difflog_ger, order = c(1, 2, 1), seasonal = list(order = c(1, 1, 0), period = 12))
```

```{r}
#Canada - ARIMA Summary
cat(blue$bold('Canada : \n'))
cat(black$bold('.: SARIMA Canada Accuracy :.'))
summary(sarima_121_can)
cat(blue$bold('German : \n'))
cat(black$bold('\n.: SARIMA German Accuracy :.'))
summary(sarima_121_ger)

```

**Forecasting using SARIMA**

```{r}

#Plot SARIMA Model
plot(forecast(sarima_121_can, h = 10), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(sarima_121_can), col = 'goldenrod2', lwd = 2.5)

#Plot
grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - Canada', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using  SARIMA Model - SARIMA(1, 2, 1)(1, 1, 0)_{12}', cex.main = 1, font.main = 3, line = 0.2)


#Forecast
cat(black$bold('.: Canada - Forecast Results (SARIMA) :.\n'))
forecast(sarima_121_can, h = 10)
cat(reset('\n'))
```
```{r}
#Plot
plot(forecast(sarima_121_ger, h = 10), main = '', xlab = '', ylab = '', bty = 'L')
lines(fitted(sarima_121_ger), col = 'goldenrod2', lwd = 2.5)

grid(nx = NA, ny = NULL, lty = 2, col = 'gray85', lwd = 1)
title(main = 'Forecast Results - German', xlab = 'Year', ylab = 'Values', cex.main = 1.5, cex.lab = 1.15, font.lab = 2)
title(main = 'using  SARIMA Model - SARIMA(1, 2, 1)(1, 1, 0)_{12}', cex.main = 1, font.main = 3, line = 0.2)


#Forecast 
cat(black$bold('.: German - Forecast Results (SARIMA) :.\n'))
forecast(sarima_121_ger, h = 10)
cat(reset('\n'))
```

